Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-04-17 src/tex/ms.tex
[showyourwork] overleaf sync

\newcommand{\rainbows}{%
    \textcolor{C0}{r}%
    \textcolor{C1}{a}%
    \textcolor{C2}{i}%
    \textcolor{C3}{n}%
    \textcolor{C4}{b}%
    \textcolor{C5}{o}%
    \textcolor{C6}{w}%
    \textcolor{C7}{s}%
    % \xspace
}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (HWO), a coronagraphic mission to detect rocky planets in habitable zones, establish their habitability, and search for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. 
%
Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. 
%
The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. 
%
Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. 
%
We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. 
%
We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. 
%
The glint signature at planetary phase angles of 
$\sim$50--70$^\circ$ 
%\sim$110--130$^\circ$ 
would be accessible to the Habitable Worlds Observatory in $\sim$14 systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in $\sim$42 systems, assuming a 62\,mas inner working angle (\IWA; 3$\lambda/D$ for a 6-m telescope at 600nm).
%
Increasing the \IWA to 2$\lambda/D$, increases the number of systems with detectable rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA{}.    \end{abstract}

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep{decadal} recommended the first in a new \enquote{Great Observatories} program a telescope with the capability to detect signatures of habitability on about 25 habitable zone planets.
%
This assertion requires an instrument with a coronagraph capable of high contrast imaging at optical to near infrared wavelengths.
%
Following this release, NASA recently announced the start of the development of the Habitable Worlds Observatory (\hwo).
%
The precursor technology recommended by the Decadal Survey also lists ``direct imaging to probe polarized ocean glint on terrestrial planets'' as one of the priority capabilities \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
The performance and precise characteristics of this telescope (on-axis or off-axis, exact diameter, type of segmentation, type of coronagraph) are still to be determined.
%
The development will be heavily influenced by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
The Inner Working angle (\IWA) of the resulting telescope will depend on these choices, which in turn will have a significant impact on the expected exo-Earth yield \citep{Stark2019_exoplanetyield}.

@@

\Cref{fig:scatterplot} shows $\Delta \varphi$ for the systems in the \hwo{} target list, assuming an IWA of \SI{62}{\mas} ($3 \lambda / D$ coronagraph at $\SI{600}{\nano\meter}$). 
\timmy{60 (as in the abstract), or 62?}
%
%This assumes each planet has a semi-major axis such that it receives an Earth-equivalent instellation, which is provided in the input star list.
%
For edge-on circular orbits, it will be possible to observe Rayleigh scattering for all targets and rainbows will be accessible in some cases.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/accessible_phase_angles.png}  
    \caption{
        \todo{shouldn't lambda/D for 600nm give 21, 41, 62, 83 mas?} 
        Cumulative distributions of the most extreme scattering phase angle accessible for different \IWA where the solid lines are for randomly inclined circular orbits and the dashed lines are for randomly orientated elliptical orbits.
        %
        The top x axis indicates the minimum scattering phase angle and the bottom x axis indicates the maximum.
        %
        These are symmetric about quadrature (90 degrees) \todo{is this true for elliptical orbits?}.
        %
        The y axis indicates the number of systems divided by the number of Monte Carlo samples and is thus normalised to the number of systems in the target list.
        %
        The secondary y axis indicates the number of systems assuming only 24 percent of them contain an Earth-like exoplanet in the habitable zone.
    }
    \label{fig:betaallofit}
    \script{plot_n_versus_phase.py}
\end{figure*}

\subsubsection{Eccentric randomly inclined orbits}
\label{sec:eccentric}
We repeat the Monte Carlo simulation from \cref{sec:circular} assuming the orbital eccentricity follows a beta distribution with shape parameters $a=0.867$ and $b=3.03$ \citep{2013MNRAS.434L..51K}. 
%
We randomly drew 1000 random orbits for each star and determined the sky-projected orbits following, e.g., \cite{2010exop.book...15M}. Eccentricities were drawn from the beta distribution, and inclinations were uniform in $\cos i$. \Cref{fig:ball-o-yarn} shows a sample of the generated orbits with the filtered \IWA's
%
%Each plot has been scaled such that the habitable zone occupies the same radius, hence the \IWA's vary in apparent size compared to the orbit.
%
We then calculated the minimum and maximum accessible scattering phase angles for of \IWA's of 21, 42, 63 and 84 mas which correspond to 1, 2, 3 and 4 times $\lambda / D$ at $\SI{600}{\nano\meter}$.
%
The dashed lines in \cref{fig:betaallofit} shows the normalised cumulative distribution of the accessible angles for each of the \IWA, when accounting for orbital eccentricity.
%
We also calculate which targets would be observable at the scattering phase angles where ocean glint, rainbows, or the Rayleigh peak occur.
%
The cumulative distribution over \IWA{} of the number of systems, normalised to the total number of systems, where each scattering phenomena is visible is shown in \cref{fig:accessible_phase_angles}.
%

@@

\begin{table}
    \centering
    \caption{
        Number of stars around which the peak phase angle of each phenomena is detectable, expressed in $mas$ and $\lambda/D$ for $\lambda=600$nm. 
    }
    \label{tab:nstars_detect}
    \begin{tabular}{ c c c c } 
    \toprule
      & \multicolumn{3}{c}{IWA}
    \\
     & 21$mas$ & 41$mas$ & 62$mas$
    \\
     & $2\lambda/D$ & $3\lambda/D$ & $4\lambda/D$ 
    \\
    \midrule
    \midrule
    Rainbows & 90 & 42 & 18 \\
    Rayleigh & 152 & 115 & 65 \\
    Ocean Glint & 43 & 14 & 5 \\
    % Glory & 0 & 5 & 10 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

The workshop on which this manuscript is based was made possible thanks to the logistical and financial support of the Lorentz Center, Leiden, Netherlands.
%
This workshop was supported by NOVA  and by the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement n°866001 - EXACT).
%
This research has made use of NASA's Astrophysics Data System Bibliographic Services and the SIMBAD database, operated at CDS, Strasbourg, France. 
%
SRV acknowledges funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation program under grant agreement No 805445.
%
SLC acknowledges support from an STFC Ernest Rutherford Fellowship. 
%
TDG acknowledges funding from the Max Planck ETH Center for Learning Systems.
%
Specific grant funding and any other specific requirements
%
KMB acknowledges support from NASA Habitable
%
Worlds grant No. 80NSSC20K152, and previous support for related work supported by NASA Astrobiology Institute's Virtual Planetary Laboratory under Cooperative Agreement Number NNA13AA93A.

>>>>

\newcommand{\rainbows}{% MAK: KEEP this formatting for MNRAS submission. It deserves this!
    \textcolor{C0}{r}%
    \textcolor{C1}{a}%
    \textcolor{C2}{i}%
    \textcolor{C3}{n}%
    \textcolor{C4}{b}%
    \textcolor{C5}{o}%
    \textcolor{C6}{w}%
    \textcolor{C7}{s}%
    % \xspace
}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (HWO), a coronagraphic mission to detect rocky planets in habitable zones, establish their habitability, and search for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. 
%
Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. 
%
The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. 
%
Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. 
%
We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. 
%
We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. 
%
The glint signature at planetary phase angles of $\sim$50--70$^\circ$ 
%\sim$110--130$^\circ$ 
would be accessible to the Habitable Worlds Observatory in $\sim$14 systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in $\sim$42 systems, assuming a 62\,mas inner working angle (\IWA{}; 3$\lambda/D$ for a 6-m telescope at 600nm).
%
Increasing the \IWA{} to 2$\lambda/D$, increases the number of systems with detectable rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA{}.    \end{abstract}

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep{decadal} recommended the first in a new \enquote{Great Observatories} program a telescope with the capability to detect signatures of habitability on about 25 habitable zone planets.
%
This assertion requires an instrument with a coronagraph capable of high contrast imaging at optical to near infrared wavelengths.
%
Following this release, NASA recently announced the start of the development of the Habitable Worlds Observatory (\hwo).
%
The precursor technology recommended by the Decadal Survey also lists ``direct imaging to probe polarized ocean glint on terrestrial planets'' as one of the priority capabilities \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
The performance and precise characteristics of this telescope (on-axis or off-axis, exact diameter, type of segmentation, type of coronagraph) are still to be determined.
%
The development will be heavily influenced by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
The Inner Working angle (\IWA{}) of the resulting telescope will depend on these choices, which in turn will have a significant impact on the expected exo-Earth yield \citep{Stark2019_exoplanetyield}.

@@

\Cref{fig:scatterplot} shows $\Delta \varphi$ for the systems in the \hwo{} target list, assuming an IWA of \SI{62}{\mas} ($3 \lambda / D$ coronagraph at $\SI{600}{\nano\meter}$). 
%\timmy{60 (as in the abstract), or 62?}
%
%This assumes each planet has a semi-major axis such that it receives an Earth-equivalent instellation, which is provided in the input star list.
%
For edge-on circular orbits, it will be possible to observe Rayleigh scattering for all targets and rainbows will be accessible in some cases.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/accessible_phase_angles.png}  
    \caption{
        \todo{shouldn't lambda/D for 600nm give 21, 41, 62, 83 mas?} 
        Cumulative distributions of the most extreme scattering phase angle accessible for different \IWA{} where the solid lines are for randomly inclined circular orbits and the dashed lines are for randomly orientated elliptical orbits.
        %
        The top x axis indicates the minimum scattering phase angle and the bottom x axis indicates the maximum.
        %
        These are symmetric about quadrature (90 degrees) \todo{is this true for elliptical orbits?}.
        %
        The y axis indicates the number of systems divided by the number of Monte Carlo samples and is thus normalised to the number of systems in the target list.
        %
        The secondary y axis indicates the number of systems assuming only 24 percent of them contain an Earth-like exoplanet in the habitable zone.
    }
    \label{fig:betaallofit}
    \script{plot_n_versus_phase.py}
\end{figure*}

\subsubsection{Eccentric randomly inclined orbits}
\label{sec:eccentric}
We repeat the Monte Carlo simulation from \cref{sec:circular} assuming the orbital eccentricity follows a beta distribution with shape parameters $a=0.867$ and $b=3.03$ \citep{2013MNRAS.434L..51K}. 
%
We randomly drew 1000 random orbits for each star and determined the sky-projected orbits following, e.g., \cite{2010exop.book...15M}. Eccentricities were drawn from the beta distribution, and inclinations were uniform in $\cos i$. \Cref{fig:ball-o-yarn} shows a sample of the generated orbits with the filtered \IWA{}'s
%
%Each plot has been scaled such that the habitable zone occupies the same radius, hence the \IWA's vary in apparent size compared to the orbit.
%
We then calculated the minimum and maximum accessible scattering phase angles for of \IWA{}'s of 21, 42, 63 and 84 mas which correspond to 1, 2, 3 and 4 times $\lambda / D$ at $\SI{600}{\nano\meter}$.
%
The dashed lines in \cref{fig:betaallofit} shows the normalised cumulative distribution of the accessible angles for each of the \IWA, when accounting for orbital eccentricity.
%
We also calculate which targets would be observable at the scattering phase angles where ocean glint, rainbows, or the Rayleigh peak occur.
%
The cumulative distribution over \IWA{} of the number of systems, normalised to the total number of systems, where each scattering phenomena is visible is shown in \cref{fig:accessible_phase_angles}.
%

@@

\begin{table}
    \centering
    \caption{
        Number of stars around which the peak phase angle of each phenomena is detectable, expressed in $mas$ and $\lambda/D$ for $\lambda=600$nm. 
    }
    \label{tab:nstars_detect}
    \begin{tabular}{ c c c c } 
    \toprule
      & \multicolumn{3}{c}{IWA}
    \\
     & 21\,$mas$ & 41\,$mas$ & 62\,$mas$
    \\
     & $2\lambda/D$ & $3\lambda/D$ & $4\lambda/D$ 
    \\
    \midrule
    \midrule
    Rainbows & 90 & 42 & 18 \\
    Rayleigh & 152 & 115 & 65 \\
    Ocean Glint & 43 & 14 & 5 \\
    % Glory & 0 & 5 & 10 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

The workshop on which this manuscript is based was made possible thanks to the logistical and financial support of the Lorentz Center, Leiden, Netherlands.
%
This workshop was supported by NOVA  and by the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement n°866001 - EXACT).
%
This research has made use of NASA's Astrophysics Data System Bibliographic Services and the SIMBAD database, operated at CDS, Strasbourg, France. 
%
SRV acknowledges funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation program under grant agreement No 805445.
%
SLC acknowledges support from an STFC Ernest Rutherford Fellowship. 
%
TDG acknowledges funding from the Max Planck ETH Center for Learning Systems.
%
Specific grant funding and any other specific requirements
%
KMB acknowledges support from NASA Habitable
%
Worlds grant No. 80NSSC20K152, and previous support for related work supported by NASA Astrobiology Institute's Virtual Planetary Laboratory under Cooperative Agreement Number NNA13AA93A.
%
This paper has been formatted in the {\tt showyourwork!} framework, which includes the computer code to reproduce the figures, including the associated data, within an online repository to make it reproducible \citep{Luger2021}.
